The Gillespie Algorithm: Stochastic Modeling of Molecular Reaction Networks

### 2024-11-27 paper.Rmd
fixed formatting of references and appendix
updated for uniform formatting

Therefore, this simulation involves 4 species and 3 reactions channels: pyrimidine monomers Thymine 1 (T1), Thymine 2 (T2), and dimer photoproducts CPD and 64-PP. 
\[
T1 + T2 \xrightarrow{c1} CPD
\]
\[
T1 + T2 \xrightarrow{c2} 64PP
\]

@@

```{r fig-dimerization-sim, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Thymine Photodimerization Reaction Set Simulation \\label{fig:dimerization-sim}"}
# Reshape data for plotting
simulation_long <- melt(simulation, id.vars = "Time")
# Plot results
ggplot(simulation_long, aes(x = Time, y = value, color = variable)) +
  geom_line() +
  labs(
    x = "Time",
    y = "Molecule Count",
    color = "Species"
  ) +
  theme_minimal()
```

@@

Both CPD and 64-PP formation follow logarithmic growth, as expected under specific UV radiation according to previously reported quantum yields (@MartinezFernandez2022; @Mouret2006). The number of reactions per unit time plateaus as the pyrimidine pairs decrease, consistent with the propensity functions.
Due to the stochastic nature of molecular reactions, this algorithm reflects fluctuations of photodimerization as it models the probabilistic process of each reaction (Figure \ref{fig:dimerization-sim}).

### Biological Significance
- **Dimerization**
As pyrimidine photodimerization transforms two pyrimidine monomers into a dimer, this simulation has accurately represented the formation of two possible dimers, CPD and 64-PP, from pairs of thymine monomers. Dimerization is also known to occur between pairs of cytosine, as well as between thymine and cytosine, with varying rates based on the monomers involved as well as the level of UV radiation (@Mouret2006).
- **Real-World Application**
Dimerization is widely studied, as it can lead to DNA mutations causing skin cancer.The Gillespie  algorithm allows us to realistically model these reactions, as it allows us to demonstrate that our parameters of interest change over time.
This algorithm can be further adjusted to show molecular reactions with various molecular species and reaction channels, with applications in biology, chemistry, and bioinformatics.

@@

# References

@@

# Appendix

## R Code for [4.1 Photodimerization Simulation](#sec-Photodimerization)

>>>>

Therefore, this simulation involves four species: pyrimidine monomers Thymine 1 (\( T1 \)), Thymine 2 (\( T2 \)), and dimer photoproducts \( CPD \) and \( 64PP \); and two reactions channels: \( c1 \) and \( c2 \).
\[
T1 + T2 \xrightarrow{c1} CPD
\]
\[
T1 + T2 \xrightarrow{c2} 64PP
\]

@@

```{r fig-dimerization-sim, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Thymine Photodimerization Reaction Set Simulation \\label{fig:dimerization}"}
# Reshape data for plotting
simulation_long <- melt(simulation, id.vars = "Time")
# Plot results
ggplot(simulation_long, aes(x = Time, y = value, color = variable)) +
  geom_line() +
  labs(
    x = "Time",
    y = "Molecule Count",
    color = "Species"
  ) +
  theme_minimal()
```

@@

As shown in Figure \ref{fig:dimerization}, both CPD and 64-PP formation follow logarithmic growth, as expected under specific UV radiation according to previously reported quantum yields (@MartinezFernandez2022; @Mouret2006). The number of reactions per unit time plateaus as the pyrimidine pairs decrease, consistent with the propensity functions.
Due to the stochastic nature of molecular reactions, this algorithm reflects fluctuations of photodimerization as it models the probabilistic process of each reaction.

#### Dimerization

As pyrimidine photodimerization transforms two pyrimidine monomers into a dimer, this simulation has accurately represented the formation of two possible dimers, CPD and 64-PP, from pairs of thymine monomers. Dimerization is also known to occur between pairs of cytosine, as well as between thymine and cytosine, with varying rates based on the monomers involved as well as the level of UV radiation (@Mouret2006).

#### Implications and Real-World Applications
Dimerization is widely studied, as it can lead to DNA mutations causing skin cancer.The Gillespie  algorithm allows us to realistically model these reactions, as it allows us to demonstrate that our parameters of interest change over time, rather than following a general logarithmic curve.This algorithm can be further adjusted to show molecular reactions with various molecular species and reaction channels. 
Using the Gillespie algorithm to model molecular reactions, such as the dimerization of thymine shown in this simulation, has wide-ranging implications in fields such as genetics, cancer research, pharmacology, environmental science, and biotechnology. By capturing the inherent randomness in molecular processes, this approach can provide valuable insights into complex biological systems, enabling better prediction of outcomes in disease development, therapy, and environmental impact studies.

@@

\newpage
# References
<div id="refs"></div>
\printbibliography[heading=none]
\def\printbibliography{}

@@

\appendix
\section{Appendix}
## R Code for [4.1 Photodimerization Simulation](#sec-Photodimerization)
